First and foremost, I am deeply grateful to my advisor, Marco Gaboardi, and to Arthur Azevedo de Amorim, formerly a post-doctoral fellow and now a professor at Rochester Institute of Technology. Their tireless mentorship and profound expertise in the subject matter have been crucial to my PhD journey, not only guiding me in academic pursuits but also imparting invaluable life lessons. I also want to thank my former advisor, Hongwei Xi, whose passion for knowledge and pursuit of excellence have been a constant source of inspiration during my PhD research and studies.

I am also thankful to my group-mates and professors with in the POPV group, with whom I have engaged in numerous insightful discussions on academic topics, which significantly broadened my understanding of various subjects in computer science and programming languages. Beyond academia, you have been wonderful companions during our enjoyable gatherings.

Last but certainly not least, I extend heartfelt gratitude to my beloved family: my parents, whose unwavering support and invaluable guidance have been constants throughout my life, leaving me with enduring benefits. And my most cherished wife, Shi, whose everlasting kindness and adventurous spirit have profoundly enriched not only my PhD journey but also my life. Her presence and encouragement have been a constant source of motivation, and together, we have discovered new hobbies that have sustained and inspired me through this challenging journey.

\vskip 1in

\noindent
Cheng Zhang\\